---
title: 'Lab 5: Principal components analysis, ggplot, Shiny'
author: "Sophia Leiker"
date: "2/3/2022"
output: html_document
---

```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(here)
 
# For Part 1 (PCA)
library(ggfortify) # For PCA biplot
library(palmerpenguins)
library(patchwork)
 
# For Part 2 (ggplot customization)
library(readxl)
library(gghighlight) # For useful highlighting
library(lubridate) # For easier dates & times
library(plotly) # Interactive graphs

```

# Part 1: Principal components analysis (PCA)

Principal components analysis is an ordination method allowing us to glean as much about our multivariate data as possible in a simplified number of dimensions.

Here, we'll use the `penguins` data within the `palmerpenguins` R package to explore variable relationships and clustering by species in a PCA biplot. For this example with PCA, we will only use the structure size measurements (bill length and depth, mass, and flipper length).

```{r}
penguin_clean <- penguins %>% 
  drop_na()

#creating a pca object type (for `prcomp` we can only give it numeric data! need to get rid of things that are characters)
penguin_pca <- penguin_clean %>% 
  dplyr::select(body_mass_g, ends_with("_mm")) %>% #selecting all the columns that end with "_mm"
  scale() %>% #since things are on different scales, we need to rescale
  prcomp() #doing the PCA
```

```{r}
# See the loadings (weighting for each principal component)
penguin_pca$rotation

#to see the SD
penguin_pca$sdev
```

## Biplotby hand

```{r}
loadings_df <- data.frame(penguin_pca$rotation) %>% 
  mutate(axis = row.names(.)) #this adds a new row, the "." takes what ever comes out above from the pipe

new_pts_df <- data.frame(penguin_pca$x)

#using a ggplot to visualize
ggplot() +
  geom_point(data = new_pts_df, aes(x = PC1, y = PC2), color = 'blue') +
  geom_segment(data = loadings_df, #this creates a line segment, they all start from 0,0
               x = 0, y = 0, aes(xend = PC1, yend = PC2, group = axis),
               arrow = arrow(length = unit(.25, 'cm'))) + #adding arrows for principal components
  geom_text(data = loadings_df, #this is labeling the labels
            aes(x = PC1, y = PC2, label = axis), vjust = 0, nudge_y = .01) + #the nudge y makes the text float just a little above the point
  theme_minimal()
```

And by auto-plot

```{r}
autoplot(penguin_pca,
         data = penguin_clean,
         loadings = TRUE, #to show the arrows
         colour = "species",
         loadings.label = TRUE, #to add the labels to the arrows
         loadings.colour = "black",
         loadings.label.colour = "black",
         loadings.label.vjust = -0.5) +
  scale_color_manual(values = c("blue","purple","orange")) +
  scale_fill_manual(values = c("blue","purple","orange")) +
  theme_minimal()

```

## Screenplot by hand

First by hand, then by screenplot

```{r}
sd_vec <- penguin_pca$sdev
var_vec <- sd_vec^2 # Standard deviation is sqrt of variance!

pc_names <- colnames(penguin_pca$rotation) #PC names will create names

#percent explained dataframe
pct_expl_df <- data.frame(v = var_vec, #variance vector
                          pct_v = var_vec / sum(var_vec), #percent variance (each value divided by the sum of variance vector)
                          pc = fct_inorder(pc_names)) %>% #puts it in order of pc_names
  mutate(pct_lbl = paste0(round(pct_v*100,1), "%"))

ggplot(pct_expl_df, aes(x = pc, y = v)) + #mapping principal components on the x, and on y we are putting the variance (putting percent variance as a label on top)
  geom_col() +
  geom_text(aes(label = pct_lbl), vjust = 0, nudge_y = .002) + #adding the percent variance label on top (percent of total varience explained by each Principal component), moving the labels up just a bit 
  labs(x = 'Principal component', y = 'Variance explained')

```

### Using function `screeplot` to do the exact same as above

- It is much harder to customize 

```{r}
# Variance explained by each PC
screeplot(penguin_pca, type = "lines")
screeplot(penguin_pca, type = "barplot")
```

