---
title: 'Lab 5: Principal components analysis, ggplot, Shiny'
author: "Sophia Leiker"
date: "2/3/2022"
output: html_document
---

```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(here)
 
# For Part 1 (PCA)
library(ggfortify) # For PCA biplot
library(palmerpenguins)
library(patchwork)
 
# For Part 2 (ggplot customization)
library(readxl)
library(gghighlight) # For useful highlighting
library(lubridate) # For easier dates & times
library(plotly) # Interactive graphs

```

# Part 1: Principal components analysis (PCA)

Principal components analysis is an ordination method allowing us to glean as much about our multivariate data as possible in a simplified number of dimensions.

Here, we'll use the `penguins` data within the `palmerpenguins` R package to explore variable relationships and clustering by species in a PCA biplot. For this example with PCA, we will only use the structure size measurements (bill length and depth, mass, and flipper length).

```{r}
penguin_clean <- penguins %>% 
  drop_na()

#creating a pca object type (for `prcomp` we can only give it numeric data! need to get rid of things that are characters)
penguin_pca <- penguin_clean %>% 
  dplyr::select(body_mass_g, ends_with("_mm")) %>% #selecting all the columns that end with "_mm"
  scale() %>% #since things are on different scales, we need to rescale
  prcomp() #doing the PCA
```

```{r}
# See the loadings (weighting for each principal component)
penguin_pca$rotation

#to see the SD
penguin_pca$sdev
```

## Biplotby hand

```{r}
loadings_df <- data.frame(penguin_pca$rotation) %>% 
  mutate(axis = row.names(.)) #this adds a new row, the "." takes what ever comes out above from the pipe

new_pts_df <- data.frame(penguin_pca$x)

#using a ggplot to visualize
ggplot() +
  geom_point(data = new_pts_df, aes(x = PC1, y = PC2), color = 'blue') +
  geom_segment(data = loadings_df, #this creates a line segment, they all start from 0,0
               x = 0, y = 0, aes(xend = PC1, yend = PC2, group = axis),
               arrow = arrow(length = unit(.25, 'cm'))) + #adding arrows for principal components
  geom_text(data = loadings_df, #this is labeling the labels
            aes(x = PC1, y = PC2, label = axis), vjust = 0, nudge_y = .01) + #the nudge y makes the text float just a little above the point
  theme_minimal()
```

And by auto-plot

```{r}
autoplot(penguin_pca,
         data = penguin_clean,
         loadings = TRUE, #to show the arrows
         colour = "species",
         loadings.label = TRUE, #to add the labels to the arrows
         loadings.colour = "black",
         loadings.label.colour = "black",
         loadings.label.vjust = -0.5) +
  scale_color_manual(values = c("blue","purple","orange")) +
  scale_fill_manual(values = c("blue","purple","orange")) +
  theme_minimal()

```

## Screenplot by hand

First by hand, then by screenplot
Note: can use summary(penguin_pca) to get a breakdown of SD, proportion of varience, cumulative proportion

```{r}
sd_vec <- penguin_pca$sdev
var_vec <- sd_vec^2 # Standard deviation is sqrt of variance!

pc_names <- colnames(penguin_pca$rotation) #PC names will create names

#percent explained dataframe
pct_expl_df <- data.frame(v = var_vec, #variance vector
                          pct_v = var_vec / sum(var_vec), #percent variance (each value divided by the sum of variance vector)
                          pc = fct_inorder(pc_names)) %>% #puts it in order of pc_names
  mutate(pct_lbl = paste0(round(pct_v*100,1), "%"))

ggplot(pct_expl_df, aes(x = pc, y = v)) + #mapping principal components on the x, and on y we are putting the variance (putting percent variance as a label on top)
  geom_col() +
  geom_text(aes(label = pct_lbl), vjust = 0, nudge_y = .002) + #adding the percent variance label on top (percent of total varience explained by each Principal component), moving the labels up just a bit 
  labs(x = 'Principal component', y = 'Variance explained')

```

With these two top principal components we have a pretty good understanding of what variables should be used. (Tells us the importance of the variables and the ways these variables relate to one another). So this helps us make more informed decisions about what might be most important to include (variable-wise)

### Using function `screeplot` to do the exact same as above

- It is much harder to customize 

```{r}
# Variance explained by each PC
screeplot(penguin_pca, type = "lines")
screeplot(penguin_pca, type = "barplot")
```


***

# Part 2: `ggplot` customization & reading in different file types

We spent some time in ESM 206 customizing our data visualizations. Let's add some more tools, including:
- Highlight spaghetti plots with `gghighlight`
- An interactive graph with `plotly`

Here, we'll also read in stored .txt and .xlsx, and files from a URL to build our toolkit for how to read in data.

**Data:** NOAA [Foreign Fisheries Trade Data](https://www.fisheries.noaa.gov/national/sustainable-fisheries/foreign-fishery-trade-data)

## Read in a .xlsx file, & do some wrangling

```{r}
fish_noaa <- read_excel(here("data", "foss_landings.xlsx")) %>% 
  janitor::clean_names() %>% 
  mutate(across(where(is.character), tolower)) %>% # convert all characters to lowercase, goes across columns run a test to see if they are characters and apply the function `tolower` where is.character is true
  mutate(nmfs_name = str_sub(nmfs_name, end = -4)) %>%  # remove last 4 characters
filter(confidentiality == "public")
```

Now, let's make and customize a graph:

```{r}
fish_plot <- ggplot(data = fish_noaa, aes(x = year, y = pounds, group = nmfs_name)) +
  geom_line(aes(color = nmfs_name)) +
  theme_minimal()
 
# Make it interactive:
ggplotly(fish_plot)
 
# Highlight series based on condition(s): USING `highlight`
ggplot(data = fish_noaa, aes(x = year, y = pounds, group = nmfs_name)) +
  geom_line() +
  gghighlight(nmfs_name == "rockfishes") + # Highlight just rockfishes
  theme_minimal()

# Highlight series based on high-catch
ggplot(data = fish_noaa, aes(x = year, y = pounds, group = nmfs_name)) +
  geom_line(aes(color = nmfs_name)) +
  gghighlight(max(pounds) > 1e8) + # Highlight max catch
  theme_minimal()

```

## Read in data from a URL, `lubridate()` refresher, `mutate()` as a workhorse to transform variables

- See paletteer color palettes:
- Discrete with `View(palettes_d_names)`
- Continuous with `View(palettes_c_names)`

**Data:** Monroe Water Treatment Plant Daily Electricity Use

Accessed from [data.gov](https://catalog.data.gov/dataset/monroe-water-treatment-plant-energy-and-water-consumption/resource/5afc8aa7-b485-4173-bcc6-c56270efedb8)

Archived here: 

* https://raw.githubusercontent.com/oharac/esm244_w22_lab5/master/data/MWTP_Elec_Daily.csv
* Can go to https://github.com/oharac/esm244_w22_lab5/blob/master/data/MWTP_Elec_Daily.csv then click the "view raw" button to get a downloadable .csv

Summary: "Daily energy use (kWh), demand (kW), and volume water treated (million gallons). 2010 through current. A second electric meter and account were added at the plant in March 2013. The usage and demand data from this meter are labeled as "Energy Use 2" and "Peak 2."

The URL to the CSV file is provided at the website above (or copy from below):

```{r}
monroe_wt <- read_csv("https://raw.githubusercontent.com/oharac/esm244_w22_lab5/master/data/MWTP_Elec_Daily.csv") %>%
  janitor::clean_names()

monroe_ts <- monroe_wt %>%
  mutate(date = lubridate::mdy(date)) %>% # Convert date to a stored date class, 
  mutate(record_month = lubridate::month(date)) %>% # Add column w/ month number (pulling month out of the date column)
  mutate(month_name = month.abb[record_month]) %>% # Add column w/ month abbreviation (pulling )
  mutate(month_name = fct_reorder(month_name, record_month)) # Make month name a factor & reorder based on values in record_month column
 
ggplot(data = monroe_ts, aes(x = month_name, y = total_kwh)) +
  geom_jitter(aes(color = month_name),
          	show.legend = FALSE,
          	alpha = 0.5,
          	size = 0.3,
          	width = 0.2) +
  theme_classic()
```

